---
title: "Data Exploration"
output: pdf_document
date: "2025-04-02"
---

#NOTE:
This file contains many and multiple rough drafts of plots, as well as plots that did not make it on to my report/presentation.



#List of Color Codes to Use Throughout Project

Dont forget to add #, had to remove for RMD syntax conflict

74add1
4575b4
d73027
a6cee3 : a lighter version of 74add1 for higher contrast in some scenarios



Code for manual color fill

```{r}

#scale_fill_manual(
 #   name = NULL,  
  #  values = c(
   #   "No Sleep Disorder" = "#a6cee3",
    #  "Sleep Apnea" = "#4575b4",
     # "Insomnia" = "#d73027"
   # ) 
```


Code for titles/axes and other reusable chunks

```{r}
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )




theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.key.height = unit(1.5, "lines"),
    legend.spacing.y = unit(0.5, "lines")
  )




panel.grid.minor = element_blank() +scale_x_continuous(breaks = seq(5.5, 8.5, by = 0.5))


legend.position = "none"

```



#Packages and Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
library(reshape2) 
library(plotly)
df <- read_csv("~/Desktop/Data Visualization/Sleep_health_and_lifestyle_dataset.csv")
```



#Selecting a dataset

```{r}
head(df)
dim(df)
summary(df)
sum(is.na(df))
colSums(is.na(df))
summary(df[, sapply(df, is.numeric)])
sapply(df, function(x) length(unique(x)))

colnames(df)




```

#DISTRIBUTION OF A SINGLE CATEGORICAL VARIABLE


## Showing the distribution of sleep disorders in patients in this study


Key Question: How are sleep disorders distributed among patients?

```{r}
df %>%
  count(`Sleep Disorder`, name = "n") %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(`Sleep Disorder`, -n), y = n, fill = `Sleep Disorder`)) +
  geom_bar(stat = "identity", color = "black", width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")), 
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "None" = "#a6cee3",
    "Sleep Apnea" = "#4575b4",
    "Insomnia" = "#d73027"
  )) +
  labs(
    title = "Distribution of Sleep Disorders",
    x = NULL,
    y = "Number of Individuals"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 12, margin = margin(r = 10))
  )






df %>%
  mutate(Gender = factor(Gender, levels = c("Male", "Female"))) %>%  
  count(Gender, `Sleep Disorder`, name = "n") %>%
  group_by(`Sleep Disorder`) %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ggplot(aes(x = `Sleep Disorder`, y = n, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9),
           width = 0.7, color = "black") +
  geom_text(
    aes(label = paste0(round(Percent, 1), "%")),
    position = position_dodge(width = 0.9),
    vjust = -0.6,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Male" = "#4575b4", "Female" = "#d73027")) +
  labs(
    title = "Distribution of Sleep Disorders by Gender",
    x = "Sleep Disorder",
    y = NULL,
    fill = "Gender"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
  axis.title.y = element_blank(),
  axis.title.x = element_blank(),     
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.grid.major.y = element_blank()
)



```








#DISTRIBUTION OF A SINGLE QUANTITATIVE
```{r}


ggplot(df, aes(x = `Sleep Duration`)) +
  geom_histogram(binwidth = 0.2, fill = "#74add1", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Hours of Sleep",
    y = "Number of Participants"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )


  
  
  
ggplot(df, aes(x = `Sleep Duration`)) +
  geom_density(fill = "#74add1", color = "#2b83ba", alpha = 0.8, size = 1.4, adjust = 1) +
  labs(
    title = "Sleep Duration Among Participants",
    x = "Hours of Sleep",
    y = "Density"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(t = 10))
  )




```


```
















#A chart that shows the distribution of two categorical variables

Does this show the DISTRIBUTION of two? aka gender? or is it just faceted?
Keep this one?



Key Question: Are sleep disorders/lack thereof associated with gender?
```{r}
#Categorical variable 1: Gender (x-axis)
#Categorical variable 2: Sleep Disorder (fill color)


#EDITS TO MAKE
#Get rid of some of the y-axis ticks/labels, less nevessary because each chunk is labeled already
#Add sample size for male vs female?


df %>%
  count(Gender, `Sleep Disorder`) %>%
  group_by(Gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Gender, y = prop, fill = `Sleep Disorder`)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            position = position_stack(vjust = 0.5), size = 4.5, color = "white") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("None" = "#74add1", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027")
  ) +
  labs(
    title = "Distribution of Sleep Disorders by Gender",
    x = "Gender",
    y = NULL,
    fill = "Sleep Disorder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.key.height = unit(1.5, "lines"),
    legend.spacing.y = unit(0.5, "lines")
  )

```





#A chart that shows the distribution of a quantitative variable across categories of a categorical variable

Need to pick one
Boxplot: More interesting with outliers
Violin: More interesting with distribution shape. Any way to add outliers here?

Split by gender/bmi/etc

```{r}



ggplot(df, aes(x = `Sleep Disorder`, y = `Sleep Duration`, fill = `Sleep Disorder`)) +
  geom_boxplot(
    aes(color = NULL), 
    color = "black", 
    outlier.shape = 16, 
    outlier.size = 2
  ) +
  labs(
    title = "Distribution of Sleep Duration by Sleep Disorder",
    y = "Hours of Sleep"
  ) +
  scale_y_continuous(breaks = seq(5, 9, 0.5)) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_blank()
  ) +
  scale_fill_manual(values = c("None" = "#74add1", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027")) +
  guides(fill = guide_legend(override.aes = list(color = NA)))  





```


#Two Quantitative



```{r}

#Sleep Duration by Stress Level


ggplot(df, aes(x = factor(`Stress Level`), y = `Sleep Duration`)) +
  geom_boxplot(fill = "#74add1", color = "black") +
  labs(title = "Sleep Duration by Stress Level", x = "Stress Level", y = "Sleep Duration") +
  theme_minimal(base_size = 14) +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```








One graph not in class
Maybe get rid of daily steps and other to reduce some of this noise?
```{r}

num_vars <- df[, c("Sleep Duration", "Stress Level", "Heart Rate", "Physical Activity Level", "Daily Steps")]
cor_matrix <- round(cor(num_vars, use = "complete.obs"), 2)
cor_long <- melt(cor_matrix)


ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 4) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac", midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Correlation Matrix of Sleep & Health Variables",
    x = NULL, y = NULL, fill = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )


```








Plots I made I didn't like/set aside

```{r}

ggplot(df, aes(x = `Sleep Disorder`, y = `Heart Rate`, fill = `Sleep Disorder`)) +
  geom_boxplot(alpha = 0.7, color = "black") +
  labs(
    title = "Heart Rate by Sleep Disorder",
    x = "Sleep Disorder",
    y = "Heart Rate (bpm)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )


  
  ggplot(df, aes(x = `Stress Level`, y = `Sleep Duration`)) +
  geom_point(alpha = 0.6, color = "lightblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Stress Level vs Sleep Duration",
       x = "Stress Level", y = "Sleep Duration (Hours)") +
  theme_minimal()


ggplot(df, aes(x = `Sleep Duration`, fill = `Sleep Disorder`)) +
  geom_histogram(position = "identity", alpha = 0.9, binwidth = 0.25, color = "white") +
  scale_fill_manual(
    name = NULL,  
    values = c(
      "No Sleep Disorder" = "#a6cee3",
      "Sleep Apnea" = "#4575b4",
      "Insomnia" = "#d73027"
    )
  ) +
  labs(
    title = "Sleep Duration by Sleep Disorder",
    x = "Sleep Duration (Hours)",
    y = "Number of Individuals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.justification = "center",
    legend.text = element_text(size = 11),
    legend.box.margin = margin(b =5),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20),
    panel.grid.minor = element_blank() +scale_x_continuous(breaks = seq(5.5, 8.5, by = 0.5))
  ) 



```











Mess up kinda?

### Key Question: Does stress level impact sleep duration?

Variables: Sleep Duration & Stress Level
```{r}
#Sleep Duration by Sleep Level

#Computing medians to use as manual hover info. Cannot disable specific points, so I have to disable all and add manually.

medians <- df %>%
  group_by(`Stress Level`) %>%
  summarise(med = median(`Sleep Duration`))


# Have to compute outliers manually. Lots of GPT debugging. Auto option in Plotly is not allowing me to change the color of the outlier points, and they are defaulting to light blue that doesn't work with my graphs and I am unable to change it? Reference article : https://stackoverflow.com/questions/68332945/the-property-outliercolor-of-python-plotly-graph-objects-box-marker-is-not-worki


outliers <- df %>%
  group_by(`Stress Level`) %>%
  mutate(
    Q1 = quantile(`Sleep Duration`, 0.25),
    Q3 = quantile(`Sleep Duration`, 0.75),
    IQR = Q3 - Q1,
    is_outlier = `Sleep Duration` < (Q1 - 1.5 * IQR) | `Sleep Duration` > (Q3 + 1.5 * IQR)
  ) %>%
  filter(is_outlier)
  
df$`Stress Level` <- factor(df$`Stress Level`, ordered = TRUE, levels = 1:8)



fig <- plot_ly(
  data = df,
  x = ~as.factor(`Stress Level`),
  y = ~`Sleep Duration`,
  type = "box",
  boxpoints = FALSE,  
  hoverinfo = "skip",  
  marker = list(
    color = 'rgba(105, 179, 162, 0.5)',
    line = list(color = 'black')
  )
) %>%
  add_trace(
    data = outliers,
    x = ~as.factor(`Stress Level`),
    y = ~`Sleep Duration`,
    type = "scatter",
    mode = "markers",
    marker = list(color = "red", size = 6),
    hoverinfo = "text",
    text = ~paste("Outlier:", round(`Sleep Duration`, 2)),
    showlegend = FALSE
  ) %>%
  add_trace(
    data = medians,
    x = ~as.factor(`Stress Level`),
    y = ~med,
    type = "scatter",
    mode = "markers",
    marker = list(color = "black", size = 6),
    hoverinfo = "text",
    text = ~paste("Median:", round(med, 2)),
    showlegend = FALSE
  ) %>%
  layout(
    title = list(
      text = "Sleep Duration by Stress Level",
      x = 0.5,
      font = list(size = 20)
    ),
    xaxis = list(title = "Stress Level"),
    yaxis = list(title = "Hours of Sleep"),
    margin = list(t = 80)
  )

fig





```


Analysis: This plot aims to examine the relationship between an individual's stress level and how long they sleep. There appears to be a moderately strong inverse relationship between stress level and sleep duration. Individuals with lower stress levels appear to sleep longer.
For example, the median sleep duration for group 3 averages around 8.2 hours. On the other hand, group 8 has the lowest median sleep duration, at 6 hours even. Notably, stress levels 7 and 8 also show the tightest distributions, indicating consistently low sleep duration among individuals with higher stress. Outliers appear relatively random, but mostly at the higher end of the stress levels.






