---
title: "Data Exploration"
output: pdf_document
date: "2025-04-02"
---

#List of Color Codes to Use Throughout Project

Dont forget to add #, had to remove for RMD syntax conflict

74add1
4575b4
d73027
a6cee3 : a lighter version of 74add1 for higher contrast in some scenarios





#Packages and Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
library(reshape2) 
library(plotly)
df <- read_csv("~/Desktop/Data Visualization/Sleep_health_and_lifestyle_dataset.csv")
```



#Selecting a dataset

```{r}
head(df)
dim(df)
summary(df)
sum(is.na(df))
colSums(is.na(df))
summary(df[, sapply(df, is.numeric)])
sapply(df, function(x) length(unique(x)))

colnames(df)



```

#DISTRIBUTION OF A SINGLE CATEGORICAL VARIABLE


## Showing the distribution of sleep disorders in patients in this study


Key Question: How are sleep disorders distributed among patients?

```{r}
df %>%
  count(`Sleep Disorder`, name = "n") %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(`Sleep Disorder`, -n), y = n, fill = `Sleep Disorder`)) +
  geom_bar(stat = "identity", color = "black", width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")), 
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "None" = "#a6cee3",
    "Sleep Apnea" = "#4575b4",
    "Insomnia" = "#d73027"
  )) +
  labs(
    title = "Distribution of Sleep Disorders",
    x = NULL,
    y = "Number of Individuals"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 12, margin = margin(r = 10))
  )


#Keep this graph. 
#Are colors necessary?
#Remove
```










Is this plot wrong in practice? Since the x-axis changed from the last graph. I feel like it is useful but idk. I want to use the bottom graph as one of the graphs extended to another variable (per rubric)

Key Question: How does sleep disorder prevalence vary across BMI categories?

```{r}

plot1 <- df %>%
  mutate(`BMI Category Grouped` = case_when(
    `BMI Category` %in% c("Normal", "Normal Weight") ~ "Normal",
    TRUE ~ `BMI Category`
  )) %>%
  count(`BMI Category Grouped`, `Sleep Disorder`) %>%
  group_by(`BMI Category Grouped`) %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(`BMI Category Grouped`, -n), y = n, fill = `Sleep Disorder`)) +
  geom_bar(stat = "identity", color = "black", width = 0.8) +
  scale_fill_manual(values = c("None" = "#a6cee3", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027")) +
  labs(
    title = "Sleep Disorder by BMI Category (?)",
    x = "BMI Category",
    y = "Number of Individuals",
    fill = "Sleep Disorder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

ggplotly(plot1)
```




#DISTRIBUTION OF A SINGLE QUANTITATIVE
```{r}
#Experimenting with including levels 0,1,2 in there, although there is no entries with those values. Not sure if I should display that.
#df$`Quality of Sleep` <- factor(df$`Quality of Sleep`, levels = 1:10)


ggplot(df, aes(x = factor(`Quality of Sleep`))) +
  geom_bar(fill = "#74add1", color = "black", width = 1) +
  labs(
    title = "Distribution of Sleep Quality Scores",
    x = "Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  theme(
  plot.title = element_text(hjust = 0.5)
) 

#colnames(df)







```
















#A chart that shows the distribution of two categorical variables

Does this show the DISTRIBUTION of two? aka gender? or is it just faceted?
Keep this one?



Key Question: Are sleep disorders/lack thereof associated with gender?
```{r}
#Categorical variable 1: Gender (x-axis)
#Categorical variable 2: Sleep Disorder (fill color)


#EDITS TO MAKE
#Get rid of some of the y-axis ticks/labels, less nevessary because each chunk is labeled already


df %>%
  count(Gender, `Sleep Disorder`) %>%
  group_by(Gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Gender, y = prop, fill = `Sleep Disorder`)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            position = position_stack(vjust = 0.5), size = 4.5, color = "white") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("None" = "#74add1", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027")
  ) +
  labs(
    title = "Distribution of Sleep Disorders by Gender",
    x = "Gender",
    y = NULL,
    fill = "Sleep Disorder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.key.height = unit(1.5, "lines"),
    legend.spacing.y = unit(0.5, "lines")
  )

```





#A chart that shows the distribution of a quantitative variable across categories of a categorical variable

Need to pick one
Boxplot: More interesting with outliers
Violin: More interesting with distribution shape. Any way to add outliers here?

```{r}
ggplot(df, aes(x = `Sleep Disorder`, y = `Sleep Duration`, fill = `Sleep Disorder`)) +
  geom_boxplot(color = "black", width = 0.6, outlier.shape = 21, outlier.fill = "black", outlier.size = 2) +
  scale_fill_manual(values = c("None" = "#74add1", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027")) +
  labs(
    title = "Distribution of Sleep Duration by Sleep Disorder",
    x = "Sleep Disorder",
    y = "Sleep Duration (hours)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )



ggplot(df, aes(x = `Sleep Disorder`, y = `Sleep Duration`, fill = `Sleep Disorder`)) +
  geom_violin(trim = FALSE, color = "black") +
  labs(
    title = "Distribution of Sleep Duration by Sleep Disorder",
    x = "Sleep Disorder",
    y = "Sleep Duration (hours)"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  scale_fill_manual(values = c("None" = "#74add1", "Sleep Apnea" = "#4575b4", "Insomnia" = "#d73027"))






```


#Two Quantitative



```{r}

#Sleep Duration by Stress Level


ggplot(df, aes(x = factor(`Stress Level`), y = `Sleep Duration`)) +
  geom_boxplot(fill = "#74add1", color = "black") +
  labs(title = "Sleep Duration by Stress Level", x = "Stress Level", y = "Sleep Duration") +
  theme_minimal(base_size = 14) +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )

```








One graph not in class
Maybe get rid of daily steps and other to reduce some of this noise?
```{r}

num_vars <- df[, c("Sleep Duration", "Stress Level", "Heart Rate", "Physical Activity Level", "Daily Steps")]
cor_matrix <- round(cor(num_vars, use = "complete.obs"), 2)
cor_long <- melt(cor_matrix)


ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 4) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac", midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Correlation Matrix of Sleep & Health Variables",
    x = NULL, y = NULL, fill = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

```








Plots I made I didn't like/set aside

```{r}

ggplot(df, aes(x = `Sleep Disorder`, y = `Heart Rate`, fill = `Sleep Disorder`)) +
  geom_boxplot(alpha = 0.7, color = "black") +
  labs(
    title = "Heart Rate by Sleep Disorder",
    x = "Sleep Disorder",
    y = "Heart Rate (bpm)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )








ggplot(df, aes(x = `Physical Activity Level`, y = `Sleep Duration`)) +
  geom_jitter(color = "steelblue", alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = seq(20, 100, by = 10)) +
  scale_y_continuous(breaks = seq(5, 9, by = 1), limits = c(5.5, 9)) +
  labs(
    title = "Physical Activity vs. Sleep",
    x = "Activity Level (minutes/day)",
    y = "Hours of Sleep"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 12),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.title.x = element_text(margin = margin(t = 15)), 
    axis.title.y = element_text(margin = margin(r = 15))   
  )




```























